---
title: 'Mid-term: Hotel booking Data'
author: "Eunsun Yook"
date:  "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Classification of Hotel Booking Cancellation {.tabset .tabset-fade .tabset-pills}

### Introduction 


Hotel reservation cancellation is unavoidable but one of the major challenges in Hospitality industry because it directly affects a hotel's profit margins. If a predictive model of cancellation is available in revenue management, it allows hotel managers to develop better cancellation policy, to predict demand accurately, to take actions proactively managing man-hours and resources accordingly, and to make those canceled rooms available for walk-ins or last minute bookings.

This data is extracted from Hotel Profit Management System (PMS) and it contains booking information due to arrive between 7/1/2015 and 8/31/2017 from one resort hotel and one city hotel located in Portugal. 

In this project, my goal is to develop a prediction model for classification of bookings with and without cancellation using data mining algorithms. 

### Packages Required 


Below is the list of packages used in the project.

```{r load, results="hide", message = FALSE, warning = FALSE}
library(magrittr)   
library(tidyverse)  
library(car)        
library(knitr)      
library(psych)      
```

* Package `magrittr` is for running tidyverse without error message.
* Package `tidyverse` is used for dplyr for data manipulation and ggplot2 package for creating plots all at once.  
* Package `car` is used for identifying outliers in a boxplot.
* Package `knitr` is for displaying tables.
* Package `psych` is for creating multiple histogram.


### Data Preparation {.tabset }

The data was obtained at [tidytuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-11/readme.md) and detailed information can be found in the original publication ([Nuno Antonio, Ana de Almeida, Luis Nunes, 2019](https://www.sciencedirect.com/science/article/pii/S2352340918315191)). Since it is uncommon to find real business data, the authors intended to provide the hotel demand data to researchers and educators in revenue management, machine learning or data mining.


#### Import Data


There are 119390 observation and 32 variables in `hotels.csv` file.
```{r import}

# Importing .csv files 
hotels <- read.csv("data/hotels.csv") 
# View the structure of hotels
str(hotels)
# View the first 6 rows of hotels
head(hotels)

```


#### Change Variable Types

Some integer variables have been switched to factors for the convenience in analysis.  

```{r type,  results = "hide"}
hotels$is_canceled  <- as.factor(hotels$is_canceled)
hotels$arrival_date_year   <- as.factor(hotels$arrival_date_year )
hotels$arrival_date_week_number  <- as.factor(hotels$arrival_date_week_number )
hotels$arrival_date_day_of_month   <- as.factor(hotels$arrival_date_day_of_month)
hotels$is_repeated_guest   <- as.factor(hotels$is_repeated_guest)
hotels$previous_cancellations  <- as.factor(hotels$previous_cancellations)
hotels$previous_bookings_not_canceled <- as.factor(hotels$previous_bookings_not_canceled)
hotels$previous_cancellations  <- as.factor(hotels$previous_cancellations)
hotels$booking_changes  <- as.factor(hotels$booking_changes)

```



#### Create new variables 


New variable "arrival_date" and "stay_nights" were created using existing variables. 

```{r results = "hide" }
#create new variable: arrival_date
hotels$arrival_date <-paste(hotels$arrival_date_month, 
                            hotels$arrival_date_day_of_month,
                            hotels$arrival_date_year,sep="-")
as.Date(hotels$arrival_date, format="%B-%d-%Y")
#create new variable: stays_nights
hotels %>% mutate(stays_nights= stays_in_weekend_nights + stays_in_week_nights) -> hotels_new

```


#### Regroup a categorical variable: meal


Regarding type of meal, there are 10,650 records of SC (no meal package) and 1,169 records of Undefined. The level of "Undefined" were combined as the level of SC.  

```{r }
summary(hotels_new)
hotels_new$meal[hotels_new$meal == "Undefined"] <- "SC" #levels(hotels_new$meal)[5] <- "SC"
```





#### Missing Values

There are almost no missing values except "children".

```{r}
colSums(is.na(hotels_new))

```


There are 4 missing values in "children" and they were replaced by zero.

```{r}
hotels_new$children[is.na(hotels_new$children)]  <- 0
```


#### Outliers

To see the distribution of adr (Average Daily Rate) and to investigate any existing outliers, a histogram and a boxplot were created.

```{r}
hist(hotels_new$adr)
boxplot(hotels_new$adr) 
```

The outlier in adr was identified.

```{r  results="hide"}
Boxplot(~adr, data=hotels_new, id=list(n=Inf)) 

```


The outlier was replaced by NA and below is new boxplot without the outlier.

```{r}
hotels_new[48516, 28] <- NA # row num.  is 48516; column num. for adr is 28
boxplot(hotels_new$adr)
```


#### Final data and Numerical Summary

Below is the table of variable description.

```{r warning = FALSE }

hotels_var_names <- colnames(hotels_new)
hotels_var_type <- lapply(hotels_new, class)
hotels_var_desc <- c("Hotel (City Hotel or Resort Hotel)",
                     "Value indicating if the booking was canceled (1) or not (0)",
                     "Number of days that elapsed between the entering date of the booking into the PMS and the arrival date",
                     "Year of arrival date",
                     "Month of arrival date",
                     "Week number of year for arrival date",
                     "Day of arrival date",
                     "Number of weekend nights (Saturday or Sunday) the guest stayed or booked to stay at the hotel",
                     "Number of week nights (Monday to Friday) the guest stayed or booked to stay at the hotel",
                     "Number of adults",
                     "Number of children",
                     "Number of babies",
                     "Type of meal booked. SC - no meal package; BB - Bed & Breakfast; HB - Half board (breakfast and one other meal – usually dinner); FB - Full board (breakfast, lunch and dinner)",
                     "Country of origin. Categories are represented in the ISO 3155–3:2013 format",
                     "Market segment designation. In categories, the term “TA” means “Travel Agents” and “TO” means “Tour Operators”",
                     "Booking distribution channel. The term “TA” means “Travel Agents” and “TO” means “Tour Operators”",
                     "Value indicating if the booking name was from a repeated guest (1) or not (0)",
                     "Number of previous bookings that were cancelled by the customer prior to the current booking",
                     "Number of previous bookings not cancelled by the customer prior to the current booking",
                     "Code of room type reserved. Code is presented instead of designation for anonymity reasons",
                     "Code for the type of room assigned to the booking. Sometimes the assigned room type differs from the reserved room type due to hotel operation reasons (e.g. overbooking) or by customer request. Code is presented instead of designation for anonymity reasons",
                     "Number of changes/amendments made to the booking from the moment the booking was entered on the PMS until the moment of check-in or cancellation",
                     "Indication on if the customer made a deposit to guarantee the booking. This variable can assume three categories: No Deposit – no deposit was made; Non Refund – a deposit was made in the value of the total stay cost; Refundable – a deposit was made with a value under the total cost of stay.",
                     "ID of the travel agency that made the booking",
                     "ID of the company/entity that made the booking or responsible for paying the booking. ID is presented instead of designation for anonymity reasons",
                     "Number of days the booking was in the waiting list before it was confirmed to the customer",
                     "Type of booking,  assuming one of four categories: Contract - when the booking has an allotment or other type of contract associated to it; Group – when the booking is associated to a group; Transient – when the booking is not part of a group or contract, and is not associated to other transient booking; Transient-party – when the booking is transient, but is associated to at least other transient booking",
                     "Average Daily Rate as defined by dividing the sum of all lodging transactions by the total number of staying nights",
                     "Number of car parking spaces required by the customer",
                     "Number of special requests made by the customer (e.g. twin bed or high floor)",
                     "Reservation last status, assuming one of three categories: Canceled – booking was canceled by the customer; Check-Out – customer has checked in but already departed; No-Show – customer did not check-in and did inform the hotel of the reason why",
                     "Date at which the last status was set. This variable can be used in conjunction with the ReservationStatus to understand when was the booking canceled or when did the customer checked-out of the hotel",
                     "Arrival Date",
                     "Stays in nights"
)

data_desc_df <- as_data_frame(cbind(hotels_var_names, hotels_var_type, hotels_var_desc))
colnames(data_desc_df) <-c("Variable", "Type", "Description")
kable(data_desc_df)

```


Below is the numerical summary for all variables in hotel_new.

```{r}
hotels_new %>% summary()
```




### Exploratory Data Analysis 

Below are the multiple histograms of numeric columns.

```{r}
# Plot histograms of numeric columns
multi.hist(hotels_new[,sapply(hotels_new, is.numeric)])


```

The canceled and not canceled groups were displayed for the two hotels.  


```{r}
# hotel and cancellation
hotels_new %>% ggplot( aes(x=hotel, 
                           fill=is_canceled)) + geom_bar() +
                      scale_fill_discrete(labels=c("not canceled", "canceled")) +
                     guides(fill=guide_legend(title=NULL))

```

Within each customer_type, the canceled and not canceled groups were displayed for the two hotels.
```{r}
# hotel and cancellation - facet: customer_type
hotels_new %>% ggplot( aes(x=hotel, 
                           fill=is_canceled)) + geom_bar() +
  scale_fill_discrete(labels=c("not canceled", "canceled")) +
  guides(fill=guide_legend(title=NULL)) +
  facet_grid(. ~ customer_type) + 
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

```

Within each deposit_type, the canceled and not canceled groups were displayed for the two hotels.
```{r}
# hotel and cancellation - facet: deposit_type
hotels_new %>% ggplot( aes(x=hotel, 
                           fill=is_canceled)) + geom_bar() +
  scale_fill_discrete(labels=c("not canceled", "canceled")) +
  guides(fill=guide_legend(title=NULL)) +
  facet_grid(. ~ deposit_type) + 
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
  
```

Within each market_segment, the canceled and not canceled groups were displayed for the two hotels.
```{r}
# hotel and cancellation - facet: market_segment
hotels_new %>% ggplot(aes(x=hotel, 
                           fill=is_canceled)) + geom_bar() +
  scale_fill_discrete(labels=c("not canceled", "canceled")) +
  guides(fill=guide_legend(title=NULL)) +
  facet_grid(. ~ market_segment) + 
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

```


Monthly lead_time were displayed for the canceled and not canceled groups.
```{r}
#Lead_time for canceled and not canceled group
hotels_new$arrival_date_month<- factor(hotels_new$arrival_date_month, levels = 
                                         c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))


hotels_new %>% ggplot( aes(x=arrival_date_month, 
                           y=lead_time, fill=is_canceled))+
  geom_bar(stat="identity", position="dodge") + 
  scale_fill_discrete(labels=c("not canceled", "canceled")) +
  guides(fill=guide_legend(title=NULL)) +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

```

Below are scatter plots between two numerical variables.
```{r warning = FALSE  }
# scatter plot between adr and lead_time
hotels_new %>% ggplot(aes(x=adr, y=lead_time)) +geom_point()
# scatter plot between adr and stays_nights
hotels_new %>% ggplot(aes(x=adr, y=stays_nights)) +geom_point()
```


